---
title: "DC Crimes time series forecast"
output:
  word_document: default
  html_notebook: default
---

Based on [Forecasting Chicago Crime Numbers (2017-2021) on Kaggle](https://www.kaggle.com/umeshnarayanappa/d/currie32/crimes-in-chicago/forecasting-chicago-crimes-2017-2020)

# Preparation  
Load some packges

```{r, echo=FALSE}
library(tidyverse)
library(prophet)
library(lubridate)
```


## Load the data
Right now it's four years of Crimes in the Washington DC area. 
```{r}
crime.data <- read_csv("data/dc-crime20150101-20170328.csv")
crime.data <- crime.data %>% bind_rows(read_csv("data/dc-crime20130101-20141231.csv"))

```

## Some cleaning and converting

Convert character dates to datetime

```{r}
crime.data$REPORT_DAT <- mdy_hms( crime.data$REPORT_DAT)
crime.data$START_DATE <- mdy_hms(crime.data$START_DATE)
crime.data$END_DATE <- mdy_hms(crime.data$END_DATE)

```

Get rid of reports that started a long time before the report date  
```{r}
crime.data <- crime.data %>% filter(START_DATE >= ymd("2013-01-01"))


```

## Grouping by day, month, year  

```{r}
crime.data$REPORT_DAY <- date(crime.data$REPORT_DAT)
crime.data$START_DAY <- date(crime.data$START_DATE) 
crime.data$END_DAY <- date(crime.data$END_DATE)
crime.data$START_MONTH <- floor_date(crime.data$START_DATE, unit="month")
crime.data$START_WEEK <- floor_date(crime.data$START_DATE, unit="week")



crimes.by.day <- crime.data %>% group_by(START_DAY) %>% summarise(TOTAL=n())
crimes.by.week <- crime.data %>% group_by(START_WEEK) %>% summarise(TOTAL=n())
crimes.by.month <- crime.data %>% group_by(START_MONTH) %>% summarise(TOTAL=n())


```

Just for fun, let's plot the crimes by day
```{r}
ggplot(data=crimes.by.day, aes(x=START_DAY, y=TOTAL)) +geom_line()
ggplot(data=crimes.by.week, aes(x=START_WEEK, y=TOTAL)) +geom_line()
ggplot(data=crimes.by.month, aes(x=START_MONTH, y=TOTAL)) +geom_line()


```

You see a seasonal pattern, but nothing too crazy. 

# Modeling with prophet  

## Set up the data frame for modeling  
The prophet package expects the date column to be named "ds" and the observation column "y"  
```{r}
colnames(crimes.by.day)[1:2] <- c("ds","y")

```

## Train the model  
```{r}
model <- prophet(crimes.by.day)
```



### Predict  

Make a blank data frame of future dates, one year out
```{r}
future <- make_future_dataframe(model, periods=365)
tail(future)
```

Make a forecast
```{r}
crime.forecast <- predict(model,future)
tail(crime.forecast[c('ds','yhat','yhat_lower', 'yhat_upper')])
```

plot the forecast
```{r}
plot(model, crime.forecast)
prophet_plot_components(model, crime.forecast)
```

* Overall trend is going down  
* Peaks on Friday and Saturday  
* Seasonality, probably due to weather  


### Write out to csv
```{r}
write_csv(crime.forecast, "data/crime_forecast_output.csv")

```

